Picky with peakpicking: assessing chromatographic peak quality with simple

### 2023-07-21 manuscript/manuscript.Rmd
Adding Anitra and Bryna's comments to text

Here, we manually labeled four HILIC oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics. We used these datasets to construct a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis. We found that a simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 70-80% down to 1-5% and showed minimal overfitting when applied to novel datasets. We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis and found that while only 10% of the variance in the dataset could be explained by depth in the default output from the peakpicker, approximately 40% of the variance was explained when restricted to high-quality peaks alone.

@@

Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected). Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered [@Pirttila2022; @Gloaguen2022]. However, this approach creates two downstream problems. First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of mass features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a]. Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection. Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of MFs already lost in the data collection process.

Existing peak-detection softwares do not provide a clear way to do this. Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each mass feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew [@Pirttila2022]. None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample. This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals. In an exploratory analysis, any mass feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%. Despite significant effort invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.

@@

An area particularly ripe for metabolomic analysis is that of the open ocean [@KidoSoule2015]. Low compound and high salt concentrations make this area difficult to study but its vast size and direct effect on the Earth's biogeochemistry make it critical that we understand its dynamics of energy and nutrients on a molecular scale [@Boysen2018]. Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own. These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill [@KidoSoule2015; @Boysen2021; @Heal2021].

In this paper, we use open ocean marine metabolite LC-MS samples to develop and test a variety of chromatographic peak metrics. Here, we construct and validate multiple predictive models of MF quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis. This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particulate organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.

@@

Most metrics had reasonably normal distributions after the scaling and normalization described in Methods. Visually, the most compelling separations between good and bad MFs were observed in the novel peak shape and SNR metrics, with higher peak shape correlations and larger SNR values corresponding to good MFs and almost complete separation for the peak shape (`med_cor`) metric. Peak width and its standard deviation also showed reasonable separation between good and bad MFs (good MFs tended to have low SDs and larger peak widths). The isotope shape and area correlations also showed good separation (Supp. figure 1).

@@

According to all three logistic regression models (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good. The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of MFs also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).

@@

We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, our novel peak shape correlation metric, and our novel SNR estimate, all with reported p-values < 10$^{-10}$. The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp. table 1).

The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GFF values implying that a large majority of the good MFs passed the threshold with very little noise included. The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GFF values closer to 75%. The two-parameter model performed worst when tested internally, with an FDR of about 20% and GFF also around 75% (Figure 2). However, when the models were trained on a different dataset than the one they were used to predict classifications for, they all had similar performance with FDRs around 10-25 and GFF around 60-80. The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).

@@

![Figure 2: False discovery rate (FDR) and fraction of good features found (GFF) plotted across different subsets of model parameters. Lower FDR indicates a smaller fraction of false positives among those mass features the model categorized as "Good" using a threshold of 0.5, and higher GFF indicates a larger fraction of the total good features were found using the same threshold. Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel. Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating excessive overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}

@@

stepdown_cors <- dotplot_stepdown_data %>%
  group_by(training_set, which_params) %>%
  pivot_wider(names_from = training_set, values_from = c(pred_prob)) %>%
  summarize(pcor=cor(FT2040_features, MS3000_features, method = "pearson"),
            scor=cor(FT2040_features, MS3000_features, method = "spearman")) %>% 
  mutate(pcor=paste0("r=", signif(pcor, digits = 3))) %>%
  mutate(scor=paste0("ρ=", signif(scor, digits = 3))) %>%
  mutate(which_params=factor(
    which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics")))

raw_prob_space_gp <- dotplot_stepdown_data %>%
  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(which_params=factor(
    which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
  geom_text(aes(label=pcor), x=-Inf, y=Inf, hjust=-0.1, vjust=1.5, data=stepdown_cors) +
  facet_wrap(~which_params, nrow = 1) +
  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
                    aesthetics = c("color", "fill")) +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  coord_fixed() +
  labs(x="Falkor-trained model (predicted probability)", 
       y="MESO-trained model\n(predicted probability)") +
  theme_bw() +
  theme(legend.position = "none")
rank_prob_space_gp <- dotplot_stepdown_data %>%
  group_by(training_set, which_params) %>%
  mutate(pred_prob=rank(pred_prob)) %>%
  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
  mutate(which_params=factor(
    which_params, 
    levels=c("all_params", "xcms_params", "raw_data_params"),
    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
  geom_text(aes(label=scor), x=-Inf, y=Inf, hjust=-0.1, vjust=1.5, data=stepdown_cors) +
  facet_wrap(~which_params, nrow = 1) +
  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
                    aesthetics = c("color", "fill")) +
  scale_x_continuous(breaks = c(0, nrow(MS3000_features)), 
                     labels = c("Min rank", "Max rank")) +
  scale_y_continuous(breaks = c(0, nrow(MS3000_features)),
                     labels = c("Min\nrank", "Max\nrank")) +
  labs(x="Falkor-trained model (ranked likelihood)", 
       y="MESO-trained model\n(ranked likelihood)",
       color=NULL) +
  guides(color=guide_legend(override.aes = list(size=4, alpha=0.8))) +
  coord_fixed() +
  theme_bw() +
  theme(axis.text.x = ggtext::element_markdown(hjust=c(0, 1)),
        # strip.text = element_blank(),
        strip.background = element_rect(fill="white"),
        legend.position=c(0.925, 0.25),
        legend.background = element_blank(),
        legend.spacing.x = unit(0.003, units = "npc"))
comb_prob_space_gp <- egg::ggarrange(raw_prob_space_gp, rank_prob_space_gp, draw = FALSE)
ggsave(plot = comb_prob_space_gp, 
       filename = "comb_prob_space.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

@@

None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets. All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GFF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp. figure 4). The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp. figure 4). In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.

@@

With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4). Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold. In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively. Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the GFF but expect it to be in a similar range (Figure 4).

@@

A majority of the features (1,323 of 1,832 total) in the original, non-thresholded MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 5). The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 mass features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas. The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean down to the DCM layer and decrease in concentration thereafter. A surprising number of features were also found to have DCM minima (DCM < 15m = 175m, 26 features) or linear increases with depth (15m < DCM < 175m, 12 features) given the few environmental parameters that have these trends (Figure 7).

@@

The model did fail to recover some interesting biological variation, however. Two features of particular interest were those good MFs with a DCM minimum (DCM < 15m = 175m), both of which were missed by the two-parameter model. These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker or pathway that's downregulated in excess of the biomass effect.

@@

![Figure 5: Plot of metabolite response to depth and the model classification error distribution. Barplots at the top show the number of MFs (mass features) in each depth response category and are broken down by the classification error types. Compounds were assigned a depth category via Dunn's post-hoc test for significant differences between the sample depths. Good MF found = true positive, good MF missed = false negative, bad MF found = false positive, and bad MF avoided = true negative according to a 0.5 likelihood threshold. Note that the majority of the features in the bad MF avoided category fell into the "All equal" depth class (1365 MFs) but the x-axis has been truncated at 60 for ease of visualization. The boxplots in the bottom illustrate the depth response type for 4 specific categories, with raw peak area plotted on a log scale against the sample depth (DCM = deep chlorophyll maximum, ~110 meters). All MFs are shown in the central bottom plot across three axes using the rank-normalized median value at each depth as the coordinate for that axis. Each mass feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile. Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}

@@

In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were never likely to be due entirely to chance. However, the pseudo-F and was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood (Table 1).

@@

The two metrics in the final model were rederived from the raw EIC data because they matched our intuition about what makes an MF look good to an MS expert. However, they could be implemented in XCMS directly, removing the need to revisit the raw data. It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better. For example, the method of using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly. Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research (e.g. @Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly informative parameter. This metric could be improved with more careful summary statistics that account for the differences between samples. Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few. Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms and must be recalculated, as in @Kantz2019.

@@

We were surprised at the poor performance of several other metrics. The isotope information in particular was expected to be a very strong predictor of MF quality given previous work that uses this metric extensively [@Libiseller2015; @Treutler2016; @ElAbiead2021]. We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too low-intensity to have them in this kind of dilute environmental sample.

@@

The large number of mass features due to noise present in metabolomics datasets can be controlled using a simple logistic classification model. We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model for were a custom signal-to-noise metric and a test of similarity to a bell curve. This model showed robustness to overfitting, independence from the training set, and a reduced degree of manual labeling required. With this model, we showed how the distribution of metabolites in the open ocean is strongly affected by depth and categorized molecules according to their depth response. This distribution reproduces measures of bulk biomass but highlights several molecules of interest that diverge from the overall trend.

@@

![Figure 7: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications. Plots show retention time along the x-axis in a 1 minute window around the center of the feature and show measured intensity on the y. Features are from the MESOSCOPE dataset and colored by the depth from which the biological sample was taken. DCM = deep chlorophyll maximum, approximately 110 meters. Mass feature ids are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}

@@

Distributions were visually inspected using a pairs plot and highly correlated (above a Pearson's r ~ 0.9) had one of the correlated metrics removed.

@@

The Pttime dataset was discovered on Metabolomics Workbench where it has been assigned Project ID PR001317. The data can be accessed directly via it's Project DOI: 10.21228/M8GH6P. This project dataset consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate. This dataset was chosen because of the similar LC-MS setup used as a benchmark for the performance other labs with similar setups may expect to achieve using the trained model directly.

@@

[TBD, get from Anitra. eScience?]

>>>>

Here, we manually labeled four HILIC oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics. We used these datasets to construct a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis. We implemented two novel peak quality metrics, a custom signal-to-noise metric and a test of similarity to a bell curve, both calculated from the raw data in the extracted ion chromatogram and found that these outperformed existing measurements of peak quality. A simple logistic regression model built on two metrics reduced the fraction of false positives in the analysis from 70-80% down to 1-5% and showed minimal overfitting when applied to novel datasets. We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis and found that while only 10% of the variance in the dataset could be explained by depth in the default output from the peakpicker, approximately 40% of the variance was explained when restricted to high-quality peaks alone.

@@

Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where contamination, background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected). Existing algorithms tend to favor the inclusion of false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered [@Pirttila2022; @Gloaguen2022]. However, this approach requires more time from the researcher as they manually evaluate a potentially enormous number of mass features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a]. Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the proportion of false positives (the false discovery rate or FDR) and accept that this will inherently add to the number of MFs already lost in the data collection process.

Existing peak-detection softwares do not provide a clear way to exclude false positives in an a consistent and unbiased way. Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each mass feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew [@Pirttila2022]. None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample. This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals. In an exploratory analysis, any mass feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99% likely to be real. Despite significant effort invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.

@@

An area particularly ripe for improved tools for metabolomic data analysis is that of the open ocean [@KidoSoule2015]. Low compound and high salt concentrations make metabolomics analyses difficult to study in this area but its vast size and the direct effect of its microbial communities on the Earth's biogeochemistry make it critical that we understand the transformation of energy and nutrients on a molecular scale [@Boysen2018]. Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own. These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill [@KidoSoule2015; @Boysen2021; @Heal2021].

In this paper, we use open ocean marine metabolite LC-MS samples to develop and test a variety of chromatographic peak metrics. We construct and validate multiple predictive models of MF quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis. This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particulate organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.

@@

Most metrics had reasonably normal distributions after the scaling and normalization described in Methods. Visually, the most compelling separations between good and bad MFs were observed in our peak shape and novel SNR metrics, with almost complete separation between good and bad peaks provided by the new peak shape metric alone. Peak width and its standard deviation also showed reasonable separation between good and bad MFs (good MFs tended to have low SDs and larger peak widths). The isotope shape and area correlations also showed good separation (Supp. figure 1).

@@

According to all three logistic regression models (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good. The full model (containing all evaluated peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of MFs also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).

@@

We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, our novel peak shape correlation metric, and our novel SNR estimate, all with reported p-values < 10$^{-10}$. The value of the novel parameters was then validated using a random forest model that also found them to have the highest importance (Supp. table 1).

The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR (false discovery rate, defined as the number of false positives divided by the total number of positive predictions) values in the 5-10% range and 80-90% GFF (% good features found, defined as the number of true positives divided by the total number of features manually classifed as good) values implying that a large majority of the good MFs passed the threshold with very little noise included. The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GFF values closer to 75%. The two-parameter model performed worst when tested internally, with an FDR of about 20% and GFF also around 75% (Figure 2). However, when the models were trained on a different dataset than the one they were used to predict classifications for, they all had similar performance with FDRs around 10-25 and GFF around 60-80. The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).

@@

![Figure 2: False discovery rate (FDR) and fraction of good features found (GFF) plotted across different subsets of model parameters. Lower FDR indicates a smaller fraction of false positives among those mass features the model categorized as "Good" using a threshold of 0.5, and higher GFF indicates a larger fraction of the total good features were found using the same threshold. Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel. Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}

@@

stepdown_cors <- dotplot_stepdown_data %>%
  group_by(training_set, which_params) %>%
  pivot_wider(names_from = training_set, values_from = c(pred_prob)) %>%
  summarize(pcor=cor(FT2040_features, MS3000_features, method = "pearson"),
            scor=cor(FT2040_features, MS3000_features, method = "spearman")) %>% 
  mutate(pcor=paste0("r=", signif(pcor, digits = 3))) %>%
  mutate(scor=paste0("ρ=", signif(scor, digits = 3))) %>%
  mutate(which_params=factor(
    which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
    labels=c("Full model", "XCMS-only model", "Two-parameter model")))

raw_prob_space_gp <- dotplot_stepdown_data %>%
  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(which_params=factor(
    which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
    labels=c("Full model", "XCMS-only model", "Two-parameter model"))) %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
  geom_text(aes(label=pcor), x=-Inf, y=Inf, hjust=-0.1, vjust=1.5, data=stepdown_cors) +
  facet_wrap(~which_params, nrow = 1) +
  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
                    aesthetics = c("color", "fill")) +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  coord_fixed() +
  labs(x="Falkor-trained model (predicted probability)", 
       y="MESO-trained model\n(predicted probability)") +
  theme_bw() +
  theme(legend.position = "none")
rank_prob_space_gp <- dotplot_stepdown_data %>%
  group_by(training_set, which_params) %>%
  mutate(pred_prob=rank(pred_prob)) %>%
  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
  mutate(which_params=factor(
    which_params, 
    levels=c("all_params", "xcms_params", "raw_data_params"),
    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
  geom_text(aes(label=scor), x=-Inf, y=Inf, hjust=-0.1, vjust=1.5, data=stepdown_cors) +
  facet_wrap(~which_params, nrow = 1) +
  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
                    aesthetics = c("color", "fill")) +
  scale_x_continuous(breaks = c(0, nrow(MS3000_features)), 
                     labels = c("Min rank", "Max rank")) +
  scale_y_continuous(breaks = c(0, nrow(MS3000_features)),
                     labels = c("Min\nrank", "Max\nrank")) +
  labs(x="Falkor-trained model (ranked likelihood)", 
       y="MESO-trained model\n(ranked likelihood)",
       color=NULL) +
  guides(color=guide_legend(override.aes = list(size=4, alpha=0.8))) +
  coord_fixed() +
  theme_bw() +
  theme(axis.text.x = ggtext::element_markdown(hjust=c(0, 1)),
        # strip.text = element_blank(),
        legend.position=c(0.925, 0.25),
        legend.background = element_blank(),
        legend.spacing.x = unit(0.003, units = "npc"))
comb_prob_space_gp <- egg::ggarrange(raw_prob_space_gp, rank_prob_space_gp, draw = FALSE)
ggsave(plot = comb_prob_space_gp, 
       filename = "comb_prob_space.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

@@

None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets. All three regularized regression models had similar behavior, with ridge regression (α = 0) obtaining the lowest rates for both GFF and FDR, while lasso (α = 1) obtained higher rates for both and represented a less-stringent false negative acceptance. As expected, the elastic net (α = 0.5) fell in between the two (Supp. figure 4). The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp. figure 4). In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.

@@

With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4). Of course, this low error rate meant that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold. In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively. Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the GFF but expect it to be in a similar range (Figure 4).

@@

A majority of the features (1,323 of 1,832 total) in the original, non-thresholded MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 5). The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 mass features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas. The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean down to the DCM layer and decrease in concentration at 175 meters. A surprising number of features were also found to have DCM minima (DCM < 15m = 175m, 26 features) or linear increases with depth (15m < DCM < 175m, 12 features) given the few environmental parameters that have these trends (Figure 7).

@@

The model did fail to recover some interesting biological variation, however. Two features of particular interest were those good MFs with a DCM minimum (DCM < 15m = 175m), both of which were missed by the two-parameter model. These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker that decreases despite an increase in biomass.

@@

![Figure 5: Plot of metabolite response to depth and the model classification error distribution. Barplots at the top show the number of MFs (mass features) in each depth response category and are broken down by the classification error types. Compounds were assigned a depth category via Dunn's post-hoc test for significant differences between the sample depths. Good MF found = true positive, good MF missed = false negative, bad MF found = false positive, and bad MF avoided = true negative according to a 0.5 likelihood threshold. Note that the majority of the features in the bad MF avoided category fell into the "All equal" depth class for which there was no significant differences between the depths (1365 MFs) but the x-axis has been truncated at 60 for ease of visualization. The boxplots in the bottom illustrate the depth response type for 4 specific categories, with raw peak area plotted on a log scale against the sample depth (DCM = deep chlorophyll maximum, ~110 meters). All MFs are shown in the central bottom plot across three axes using the rank-normalized median value at each depth as the coordinate for that axis. Each mass feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile. Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}

@@

In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were unlikely to be due to chance. However, the pseudo-F and was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood (Table 1).

@@

The two metrics in the final model were rederived from the raw EIC data because they matched our intuition about what makes an MF look good to an MS expert. These are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better. For example, the method of using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly. Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research (e.g. @Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly informative parameter. This metric could be improved with more careful summary statistics that account for the differences between samples. Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few. Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms and must be recalculated, as in @Kantz2019. To avoid the additional overhead of recalculation and the possiblity of raw data unavailability, we have implemented these metrics during the initial peakpicking step of XCMS in a fork of the GitHub available at https://github.com/wkumler/xcms and have submitted a pull request to implement them into XCMS directly.

@@

We were surprised at the poor performance of several other metrics. The isotope information in particular was expected to be a very strong predictor of MF quality given previous work that uses this metric extensively [@Libiseller2015; @Treutler2016; @ElAbiead2021]. We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too small (low-intensity) to have detectable isotope peaks in this kind of dilute environmental sample.

@@

The large number of mass features due to noise present in metabolomics datasets can be controlled using a simple logistic classification model. We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model were a custom signal-to-noise metric and a test of similarity to a bell curve. This model showed robustness to overfitting, independence from the training set, and a reduced degree of manual labeling required. With this model, we showed how the distribution of metabolites in the open ocean is strongly affected by depth and categorized molecules according to their depth response. This distribution reproduces measures of bulk biomass but highlights several molecules of interest that diverge from the overall trend.

@@

![Figure 7: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications. Plots show retention time along the x-axis in a 1 minute window around the center of the feature and show measured intensity on the y. Features are from the MESOSCOPE dataset and colored by the depth from which the biological sample was taken. DCM = deep chlorophyll maximum, approximately 110 meters. Mass feature identifications are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}

@@

Distributions were visually inspected using a pairs plot and highly correlated (above a Pearson's r ~ 0.9) metrics had one of the redundant metrics removed.

@@

The Pttime dataset was discovered on Metabolomics Workbench where it has been assigned Project ID PR001317. The data can be accessed directly via it's Project DOI: 10.21228/M8GH6P. This project dataset consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate. This dataset was chosen because of the similar LC-MS setup used as a benchmark for the performance that other labs with similar setups may expect to achieve using the trained model directly.

@@

This work was supported by grants from the Simons Foundation (SCOPE Award ID 329108 to AI, SF Award ID 385428 to AI).
